In this chapter we will describe a symbolic interpreter for \explanguage. First, we describe how we must extend the grammar, to be able to do symbolic execution. Next we describe the implementation of \textsl{interpExp} and \textsl{interpStm}, and how these relate to their concrete counterparts. 
\section{Extension of grammar}

The original language have three types of values, namely integers, booleans and the unit value. Symbolic execution requires that we extend the language to include symbolic counterparts of these. We therefore introduce a symbolic value which can either be a symbolic integer, a symbolic boolean or the unit value.

\begin{alignat*}{3}
	val & \in \textsc{SymbolicValue} && := \quad int_{sym} \alt
	bool_{sym} \alt unit
\end{alignat*}



 A symbolic integer can either be a concrete integer, a symbol or an arithmetic expression over these two:



\begin{alignat*}{3}
	int & \in \textsc{Integer} && := \quad 0 \alt 1 \alt -1 \alt \ldots\\
	sym & \in \textsc{Symbol} && := \quad a \alt b \alt c \alt \ldots \\	
	int_{sym} & \in \textsc{SymbolicInteger} && := \quad int \alt sym \alt int_{sym} + int_{sym} \\
	& && \quad \alt int_{sym} - int_{sym} \alt int_{sym} * int_{sym}\\
	& && \quad \alt int_{sym} / int_{sym}\\
\end{alignat*}

A symbolic boolean can either be $True$, $False$ or an symbolic boolean expression, which is a comparison of two symbolic integers. Finally a symbolic boolean can be the negation of a symbolic boolean. This extension is needed to be able to represent the two different \emph{path-constraints} that might arise from a conditional statement.

\begin{alignat*}{3}
	bool & \in \textsc{Boolean} && := \quad True \alt False\\
	bool_{sym} & \in \textsc{SymbolicBoolean} && := \quad int_{sym} < int_{sym} \alt int_{sym} > int_{sym}\\
	& && \quad \alt int_{sym} \leq int_{sym} \alt  int_{sym} \geq int_{sym}\\
	& && \quad \alt  int_{sym} == int_{sym} \alt  ! \ int_{sym}
\end{alignat*}



\section{Path-constraints}
To represent a \pc, we implement the following classes:

\begin{lstlisting}[style=simple]
case class PathConstraint(conds: List[SymbolicBool], ps: PathStatus) {
		def :+(b: SymbolicBool): PathConstraint = PathConstraint(this.conds :+ b, this.ps)
}
sealed trait PathResult
case class Certain() extends PathResult
case class Unknown() extends PathResult
\end{lstlisting}

The definition of \textsl{PathConstraint} consists of two elements. \textsl{conds} is a  list of symbolic booleans from the conditional statements met so far. The \textsl{PathStatus} tells us whether or not we can guarantee that the path constraint is in fact satisfiable. This is necessary because we allow for nonlinear constraints, which our SMT solver may fail to solve. In the case of a failure, we still explore the path but the status of the \pc will be \textsl{Unknown}, meaning that we cannot guarantee that there exists concrete input values that follow this path.

\section{Symbolic Interpretation of expressions}
To symbolically interpret an expression, we define the following function illustrated in figure \ref{interpExpSymbolic}:

\begin{figure}[!h]
	\begin{lstlisting}[style = simple]
		def interpExp(p: Prog, e: Exp, env: Map[Id, SymbolicValue], 
				pc: PathConstraint, forks: Int): 		
					List[(Result[SymbolicValue, String], PathConstraint)]
	\end{lstlisting}
	\caption{Symbolic version of \textsl{InterpExp}}	
	\label{interpExpSymbolic}
\end{figure}

\begin{figure}[!h]
	\begin{lstlisting}[style = simple]
		def interpExp(p: Prog, e: Exp, env: Map[Id, ConcreteValue]): Result[ConcreteValue, String]
	\end{lstlisting}
	\caption{Concrete version of \textsl{InterpExp}}
	\label{interpExpConcrete}
\end{figure}
\newpage

By comparing figure \ref{interpExpSymbolic} with figure \ref{interpExpConcrete}, we see that the symbolic version of \textsl{interpExp} is an extension of the version from the concrete interpreter. It still depends current environment, which is now of type \mbox{\textsl{Map[Id, SymbolicValue]}}, and the program itself. However, it now also depends on a \pc to keep track on any constraints placed on the symbolic values,
and \textsl{forks} which is the current number of times the execution has been forked. This value is used to enforce an upper bound on the number of execution paths that we explore to avoid running forever.

The return type has also changed. Instead of returning a single result, it returns a list of pairs consisting of a result and a \pc, one for each possible execution path. This also effects how we interpret the different types of expressions. Most of them consists of one or more sub-expressions, each of which may result in several different results, so we have to consider all possible combinations of results. We will demonstrate this by looking at the interpretation of arithmetic expressions and function calls. 

\subsection{Arithmetic expressions}
Consider an arithmetic expression 
\textsl{AExp(e1: Exp, e2: Exp, op: AOp)}. When we recursively interpret $e_1$ and $e_2$, we get two lists $L_{e_1}, L_{e_2}$ of possible results. We need to take the cartesian $L_{e_1} \times L_{e_2}$ of the lists, and for each pair of results, we have to evaluate the arithmetic expression. 

To do this we use a \textsl{for-comprehension} which given two lists, iterate over each ordered pair of elements from the two lists.  For each pair, we \textsl{flatMap} over the two results, and if no errors are encountered, we check that both expressions evaluates to symbolic integers and compute the result. If both sub expression evaluate to concrete integers, we compute the result as we would do in a concrete interpretation. 

\subsection{Function calls}
Consider a \textsl{Call}-expression \textsl{CallExp(id: Id, args: List[Exp])}.
First, we must check that the function is defined and that the formal and actual argument list does not differ in length. If both of these checks out, we map \textsl{interpExp} on to \textsl{args}. This gives a list of lists $[L_{e_1}, L_{e_2}, \ldots, L_{e_n}]$, where the $i'th$ list contains the possible results of interpreting the $i'th$ expression in \textsl{args}.
We take the cartesian product $L_{e_1} \times L_{e_2} \times \ldots \times L_{e_n}$, which gives us all possible argument lists. For each of these lists, we zip it with formal argument list. This gives a list of the following type \textsl{List[(ExpRes, Id)]}. We attempt to build a local environment by folding over this list, starting with the original environment that was passed with the call to \textsl{interpExp}. If we encounter an error, either from the interpretation of the expressions in \textsl{args} or from an expression evaluating to the unit value, the result of the function call with this argument list will be that error. Otherwise, for each pair of result and id, we make a new environment with the appropriate binding added. We then interpret the statement in the function body  with the local environment, and the result of the function call with this argument list, will be the value of the statement. 

\section{Symbolic interpretation of statements}

To symbolically interpret statements, we define the following function shown in figure \ref{interpStmSymbolic}:

\begin{figure}[!h]
	\begin{lstlisting}[style=simple]
		def interpStm(p: Prog, stm: Stm, env: Map[Id,SymbolicValue], 
			pc: PathConstraint, forks: Int): 
				List[(Result[SymbolicValue, String], Map[Id, SymbolicValue], 	PathConstraint)]
	\end{lstlisting}
	\caption{Symbolic version of \textsl{interpStm}}
	\label{interpStmSymbolic}
\end{figure}

\begin{figure}[!h]
	\begin{lstlisting}[style=simple]
		def interpStm(p: Prog, s: Stm, env: Map[Id, ConcreteValue]): Result[(ConcreteValue, Map[Id, ConcreteValue]), String]
	\end{lstlisting}
	\caption{Concrete version of \textsl{interpStm}}
	\label{interpStmConcrete}
\end{figure}
By comparing figure \ref{interpStmSymbolic} and figure \ref{interpStmConcrete} we see that this version of \textsl{interpStm} is also an extension of its concrete counterpart in the same way that the symbolic version of \textsl{interpExp} is. It now also depend on a \pc and the number of forked executions. Similarly the return type has changed so that it now returns a list of triples consisting of a result, a \pc and a potentially updated environment. We will now demonstrate how we interpret conditional statements. We will also see how we deal with undecidable \emph{path-constraints} and how we limit the number of paths that we explore.


\iffalse

\subsection{Assignment statements}
Given an \textsl{Assignment}-statement \textsl{AssignStm(v: Var, e: Exp)}, we interpret the expression $e$, and get a list of results for each possible execution path. For each of these results, we return the value of the expression and an updated environment. If the expression resulted in an error, or a unit 
value, we instead return error and the original environment. 

\fi

\subsection{Conditional statements}

As mentioned earlier, \explanguage supports three different types of conditional statements, which are represented by the following classes:

\begin{lstlisting}[style=simple]
	case class IfStm(cond: Exp, thenStm: Stm, elseStm: Stm)
	case class WhileStm(cond: Exp, doStm: Stm)
	case class AssertStm(cond: Exp)
\end{lstlisting}
For each possible result of interpreting the condition expression we do the following: If it is either \textsl{True()} or \textsl{False()}, the expression must have been free from symbols, so we can continue as we would have done during a concrete interpretation, with the same \pc and value of \textsl{forks}.

\noindent If the result is a symbolic boolean expression $b$, we have two potential paths to follow. We must check the satisfiability of \textsl{pc :+ b} and \textsl{pc :+ Not(b)}. We do this by calling \textsl{(checkSat(pc, b)} and \textsl{checkSat(pc, Not(b))}. If \textsl{checkSat} reports \textsl{SATISFIABLE}, the given path is eligible for exploration. If it reports \textsl{UNSATISFIABLE}, we can safely ignore the path. If it reports \textsl{UNKNOWN}, we still explore the path, but the new \pc will have its status set to \textsl{Unknown}.
 If both calls to \textsl{checksat} return either \textsl{SATISFIABLE} or \textsl{UNKNOWN}, we explore both paths, and return a list containing the results of the first path, followed by the results of the second path.

\subsubsection{Checking satisfiability}
To check the satisfiability of a path, we define the following function

\begin{lstlisting}[style=simple]
def checkSat(pc: PathConstraint, b: SymbolicBool): z3.Status
\end{lstlisting}
which takes a \pc  and a symbolic boolean $b$, and checks the satisfiability of $b_1 \land b_2 \land \ldots \land b_k \land b$, where $b_1, b_2, \ldots, b_k$ are the symbolic booleans in \textsl{pc.conds}. To do this, we use the \textsl{Java} implementation of the \textsl{Z3} SMT solver. 
 
\subsubsection{Uppper bound on number of forks}

The interpretation of an \textsl{If}-statement and a \textsl{While}-statement starts with checking whether $forks > maxForks$, in which case we immediately return an Error. We include this check to prevent the interpreter to run forever on programs with an infinite number of execution paths. 

\iffalse
\subsection{Sequence statements}

We define the following class to represent a sequence statement:
\begin{lstlisting}[style=simple]
	case class SeqStm(s1: Stm, s2: Stm)
\end{lstlisting}
For each possible result from interpreting \textsl{s1}, we interpret \textsl{s2} with the new environment. If the interpretation of \textsl{s1} results in an error, this will be the result of the sequence expression. otherwise it will be the result of the interpretation of \textsl{s2}. 	

\subsection{Expression statements}
We define the following class to represent an \textsl{expression}-statement:
\begin{lstlisting}[style=simple]
case class ExpStm(e: Exp)
\end{lstlisting}
To interpret this, we interpret $e$. For each result of this, the result of the statement, will be the value of the expression and the original environment. 

\fi